---
title: "Adult Income Analysis"
author: "Kevin E. D'Elia"
date: "October 2nd, 2016"
output:
  html_document:
    highlight: zenburn
    theme: cosmo
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path="./figures/")
```

# Executive Summary
The purpose of this analysis is to predict whether income exceeds $50K/yr based on census data. Also known as "Census Income" dataset.

# Pre-requisites/Assumptions
Before working with this dataset, it is assumed that the following pre-requisites are met:

1. The dataset **adult.data.csv** has been loaded
2. The fields are comma-separated and there is no header
3. Variable names must be taken from the data dictionary and applied to the data

# Exploring and understanding data
```{r load_training_data}
adultData <- read.table("./Adult/adult.data", sep = ",", header = FALSE)
names(adultData) <- c("age", "workclass", "fnlwgt", "education", "education_num", "marital_status", "occupation", "relationship", "race", "sex", "capital_gain", "capital_loss", "hours_per_week", "native_country", "income_level")
dim(adultData)
```

Now read in the test data, ignoring the first row (skip = 1) as it is garbage data, and remove the last column ("income_level"), since that is the column we are trying to predict:
```{r load_testing_data}
adultTest <- read.table("./Adult/adult.test", sep = ",", header = FALSE, skip = 1)
names(adultTest) <- c("age", "workclass", "fnlwgt", "education", "education_num", "marital_status", "occupation", "relationship", "race", "sex", "capital_gain", "capital_loss", "hours_per_week", "native_country", "income_level")
adultTest <- adultTest[, -15]
dim(adultTest)
```

## Exploring the structure of data
The data dictionary informs us that both datasets contain identically named variables; the datasets vary in the number of observations provided.

What is the structure of the data we are working with?
```{r structure}
str(adultData)
str(adultTest)
```

## Exploring numeric variables

The structure of the data identifies several variables that are numeric in nature; use the **summary** command to explore them:
```{r summary}
summary(adultData[c("age", "education_num", "capital_gain", "capital_loss")])
```

It looks as if the **capital_gain** variable has some bogus values; remove them and re-run statistics on just that variable:
```{r clean_up}
adultData <- adultData[which(adultData$capital_gain != 99999), ]
summary(adultData$capital_gain)
```

## Measuring the central tendency - mean and median

Measures of **central tendency** are a class of statistics used to identify a value that falls in the middle of a set of data.  What is the average, or **mean**, of the ages of our subjects?
```{r mean}
round(mean(adultData$age), 2)
```

